---
title: "Vignette of the `medicalMonitoring` package"
author: "Laure Cougnaud"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: glpgStyle::html_report
vignette: >
  %\VignetteIndexEntry{medicalMonitoring package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Introduction

The functionality to download the patient profiles require additional _Javascript_ functions:

* custom _Javascript_ functions, included in the `medicalMonitoring` package
* external _Javascript_ dependencies: _JSZip_, _JSZipUtils_

These dependencies are included once for all visualizations of the report by using:

```{r runReport, eval = FALSE}

	dep <- c(
		list(
			# downloaded from: http://github.com/Stuk/jszip/zipball/master
			htmltools::htmlDependency(
				name = "jszip",
				version = "3.2.2",
				src = c(file = "js/Stuk-jszip-3109282/dist"),
				script = "jszip.js"
			),
			# clone from: https://github.com/Stuk/jszip-utils
			htmltools::htmlDependency(
				name = "jszip-utils",
				version = "0.1.0",
				src = c(file = "js/jszip-utils/dist"),
				script = "jszip-utils.js"
			),
			# clone from: https://github.com/eligrey/FileSaver.js/
			htmltools::htmlDependency(
				name = "FileSaver",
				version = "1.3.8",
				src = c(file = "js/FileSaver.js/dist"),
				script = "FileSaver.js"
			)
		),
		getJsDepMedicalMonitoring()
	)

	# run the report
	rmarkdown::render(
		input = "medicalMonitoring-vignette.Rmd",
		output_format = rmarkdown::html_document(extra_dependencies = dep)
	)
	
```	

```{r options, echo = FALSE}
	
	library(knitr)
	opts_chunk$set(
		echo = TRUE, results = 'asis', warning = FALSE, 
		error = FALSE, message = FALSE, cache = FALSE,
		fig.width = 8, fig.height = 7,
		fig.path = "./figures_vignette/",
		#out.width = '0.7\\textwidth', 
		fig.align = 'center')#, out.height = 0.5\\textwidth', fig.path = 'graphs/') 
	options(width = 170)#, stringsAsFactors = FALSE
	options(warn = 1)#instead of warn = 0 by default -> to have place where warnings occur in the call to Sweave function
	
	heightLineIn  <- 0.2
	
```

This package `medicalMonitoring` contains visualizations/reporting functionalities
for medical monitoring.

```{r loadPackages}

	library(medicalMonitoring)

```

## Data format

The input dataset for the workflow should be a data.frame, 
usually loaded from a _SAS_ data file (`sas7bdat` format).
The label of the variables stored in the `SAS` datasets is also used
for title/captions. 

The function `loadDataADaMSDTM` can be used to load your custom `SAS` dataset(s) of
interest.
This function returns two objects:

* a list of `data.frame` containing the data for each _SAS_ dataset(s), stored
  in the `$data` slot
* a vector containing the labels for each column of the data (used e.g. for
  default titles of the figures), stored in the `$labelVars` slot

A few `sdtm` datasets from the _Pelican_ study are included in the
`glpgUtilityFct` package for the demonstration, via the dataset
`SDTMDataPelican` and corresponding variable labels `labelVarsSDTMPelican`.

```{r loadData}	
	
	library(glpgUtilityFct)
	
	data(SDTMDataPelican)
	data(labelVarsSDTMPelican)
	
	dataLB <- SDTMDataPelican$LB
	dataDM <- SDTMDataPelican$DM
	labelVars <- labelVarsSDTMPelican

```

# Data pre-processing

## Add data annotation

The `annotateData` enables to add metadata for a specific domain/dataset.

```{r annotateData}
dataLB <- annotateData(dataLB, annotations = list(data = dataDM))
```

# Visualizations

## Interactive scatterplot

### Time profiles

```{r timeProfiles}

dataPlot <- subset(dataLB, LBTESTCD == "ALT")
visitLab <- with(dataPlot, tapply(LBDY, VISIT, median))
names(visitLab) <- sub("-", "\n", names(visitLab))

# link to patient profiles
dataPlot$subjectPatientProfilePath <- paste0(
	"patientProfiles/subjectProfile-", 
	sub("/", "-", dataPlot$USUBJID), ".pdf"
)

pl <- scatterplotMonitoring(
	data = dataPlot, 
	xVar = "LBDY", yVar = "LBSTRESN",
	aesPointVar = list(color = "ACTARM", shape = "USUBJID"),
	aesLineVar = list(group = "USUBJID", color = "ACTARM"),
	hoverVar = c("USUBJID", "VISIT", "LBDY", "LBSTRESN", "COUNTRY", "ACTARM"),
	labelVars = labelVars,
	xPars = list(breaks = visitLab, labels = names(visitLab)),
	themePars = list(legend.position = "none"),
	title = paste("Actual value of", 
		getLabelParamcd(paramcd = "ALT", data = dataLB, paramcdVar = "LBTESTCD", paramVar = "LBTEST")
	),
	# include link to patient profiles:
	pathVar = "subjectPatientProfilePath"
)
pl

```

### Scatterplot

```{r scatterplot}

# format data long -> wide format (one column per lab param)
dataPlot <- subset(dataLB, LBTESTCD %in% c("ALT", "ALB"))
library(reshape2)
dataPlotWide <- dcast(
	data = dataPlot,
	formula = USUBJID + VISIT + VISITNUM ~ LBTESTCD, 
	value.var = "LBSTRESN",
	fun.aggregate = mean
)

# scatterplot per visit
scatterplotMonitoring(
	data = dataPlotWide, 
	xVar = "ALT", yVar = "ALB",
	aesPointVar = list(color = "USUBJID"),
	themePars = list(legend.position = "none"),
	facetPars = list(facets = ~ VISIT),
	labelVars = labelVars
)

```

# Appendix

## Session information

```{r includeSessionInfo, echo = FALSE}

	library(pander)
	pander(sessionInfo())

```
